---
title: "Progress Memo 1"
subtitle: |
  | Final Project 
  | Data Science 2 with R (STAT 301-2)
author: "Jessie Chen"
date: today

format:
  html:
    toc: true
    embed-resources: true
    
execute:
  echo: false
  warning: false

from: markdown+emoji 
reference-location: margin
citation-location: margin
---

::: {.callout-tip icon=false}

## Github Repo Link

To link to your github **repo**sitory, appropriately edit the example link below. Meaning replace `https://your-github-repo-url` with your github repo url. Suggest verifying the link works before submitting.

[My Repo Link](https://github.com/stat301-2-2024-winter/final-project-2-Jchench.git)

:::

## Prediction Problem

The type of prediction problem that I am focusing on for this project is categorical and I'm looking to identify whether or not a student has dropped out of college at the end of a school year (2 semesters). I think having this prediction model would be useful for this problem because it will also help indicate factors that might make a student more at risk of dropping out.

## Data source

The [dataset](https://archive.ics.uci.edu/dataset/697/predict+students+dropout+and+academic+success) that I'm using is from UC Irvine's Machine Learning Repository. This dataset was created by a number of authors, using information from higher education institution (acquired from several disjoint databases) related to students enrolled in different undergraduate degrees. ^[Realinho,Valentim, Vieira Martins,Mónica, Machado,Jorge, and Baptista,Luís. (2021). Predict students' dropout and academic success. UCI Machine Learning Repository. https://doi.org/10.24432/C5MC89.] 

*This data was mainly taken from schools in Portugal.

## Data quality & complexity check

```{r}
# load packages:
library(tidyverse)
library(here)
library(naniar)

# reading data:
drop_out_data <- 
  read_rds(here("data/dropout_data_cleaned"))
```

The dataset includes 37 variables in total: 16 categorical variables and 21 numeric variables. This dataset also has 4424 observations with each one representing a student.

```{r}
gg_miss_var(drop_out_data)
```
There are no missing variables in this dataset!

## Target variable analysis

Our target variable will be `target`, which indicates whether a student is currently enrolled, graduated, or has dropped out of college.

```{r}
drop_out_data |> 
  ggplot(aes(x = target)) +
  geom_bar()
```
We can see that the distribution is relatively imbalanced with significantly more students graduating compared to dropping out or currently enrolling. Thus, it would be helpful to use strata when splitting the data into training and learning.

## Misc

A lot of my original data was incorrectly categorized (i.e. factors were being indicated by numbers). I was able to do a bit of cleaning to change a couple of variable types. Other than that, my data is relatively clean and should be easy to work with, so I plan to have my models fitted by next week.