---
title: "Progress Memo 1"
subtitle: |
  | Final Project 
  | Data Science 2 with R (STAT 301-2)
author: "Jessie Chen"
date: today

format:
  html:
    toc: true
    embed-resources: true
    
execute:
  echo: false
  warning: false

from: markdown+emoji 
reference-location: margin
citation-location: margin
---

::: {.callout-important}
## Memo format

Like all other submissions for this course, we expect the memo to be neat and well organized. This a template to get you started. 

You should make this document your own. To help with this we have added a date to the yaml above (good whenever writing memos). You should remove this block when making this document your own.

:::

::: {.callout-tip icon=false}

## Github Repo Link

To link to your github **repo**sitory, appropriately edit the example link below. Meaning replace `https://your-github-repo-url` with your github repo url. Suggest verifying the link works before submitting.

[My Repo Link](https://github.com/stat301-2-2024-winter/final-project-2-Jchench.git)

:::

::: {.callout-warning}
GitHub (free account) cannot store large files. If you try to commit and push a large dataset you will have an ERROR! Any file over 100 MB (100,000 KB) needs to be added to the `.gitignore` file BEFORE committing.

**We may need to do that for files in your final project!**
:::

## Prediction Problem

The type of prediction problem that I am focusing on for this project is categorical and I'm looking to identify whether or not a student has dropped out of college at the end of a school year (2 semesters). I think having this prediction model would be useful for this problem because it will also help indicate factors that might make a student more at risk of dropping out.

## Data source

The [dataset](https://archive.ics.uci.edu/dataset/697/predict+students+dropout+and+academic+success) that I'm using is from UC Irvine's Machine Learning Repository. This dataset was created by a number of authors, using information from higher education institution (acquired from several disjoint databases) related to students enrolled in different undergraduate degrees. ^[Realinho,Valentim, Vieira Martins,Mónica, Machado,Jorge, and Baptista,Luís. (2021). Predict students' dropout and academic success. UCI Machine Learning Repository. https://doi.org/10.24432/C5MC89.]

## Data quality & complexity check

You should demonstrate that you can read in your dataset(s) --- since we need proof of this, showing code is permitted or show code for the next steps. Quality check means providing answers to the following: How many variables and observations do you have? How many categorical vs numerical variables? Are there any missingness issues?

```{r}
# load packages:
library(tidyverse)
library(here)
library(naniar)

# reading data:
drop_out_data <- 
  read_rds(here("data/dropout_data_cleaned"))
```

The dataset includes 37 variables in total: 16 categorical variables and 21 numeric variables. This dataset also has 4424 observations with each one representing a student.

```{r}
gg_miss_var(drop_out_data)
```
There are no missing variables in this dataset!

## Target variable analysis

A univariate analysis of the target variable. Identify and discuss potential issues (symmetry/unbalanced/skewness & missingness) and discuss potential solutions such as transformations. 
There are a lot of outliers for the target variable `loss` beyond the 75 percentile, so I placed a limit on the x-axis.

```{r}
drop_out_data |> 
  ggplot(aes(x = target)) +
  geom_bar()
```
However, we can see that the distribution is relatively unbalanced with significantly more students graducating compared to dropping out or just enrolling. Thus, it would be helpful to use strata when splitting the data into training and learning.