---
title: "Progress Memo 2"
subtitle: |
  | Final Project 
  | Data Science 2 with R (STAT 301-2)
author: "Jessie Chen"
date: today

format:
  html:
    toc: true
    embed-resources: true
    
execute:
  echo: true
  warning: false

from: markdown+emoji 
reference-location: margin
citation-location: margin
---

::: {.callout-tip icon="false"}
## Github Repo Link

[My Repo Link](https://github.com/stat301-2-2024-winter/final-project-2-Jchench.git)
:::

## Dataset and prediction problem:

The type of prediction problem that I am focusing on for this project is a **classification problem** and I'm looking to identify whether or not a student has dropped out of college at the end of a school year (2 semesters). The variable that I'm looking to predict is `target`, which is a factor with 3 levels: Dropout, Enrolled, and Graduated.

The [dataset](https://archive.ics.uci.edu/dataset/697/predict+students+dropout+and+academic+success) that I'm using is from UC Irvine's Machine Learning Repository. This dataset was created by a number of authors, using information from higher education institution (acquired from several disjoint databases) related to students enrolled in different undergraduate degrees. [^1]

[^1]: Realinho,Valentim, Vieira Martins,Mónica, Machado,Jorge, and Baptista,Luís. (2021). Predict students' dropout and academic success. UCI Machine Learning Repository. https://doi.org/10.24432/C5MC89.

## Assessment metric:

Since this is a classification problem, I will be looking at the ROC and accuracy metrics to determine which models performed better.

## Analysis plan:

Since my dataset has around 4424 observations, which is on the smaller end, I'm planning on doing a 75-25 split so that we can get more consistent and reliable estimates (3316 observations in the training dataset and 1108 observations in teh testing dataset). I'm also planning on doing vfold cross-validation to resample the training data. I will use 5 partitions (v = 5) and 10 repeats since my dataset isn't incredibly large. For this project, I plan on using 5 models: Naive Bayes (baseline model), Multinomial (because my target variable `target` has 3 factor levels), Elastic Net, K-Nearest Neighbor, Random Forest, and Boosted Tree.

## Defining Recipes:

My baseline kitchen sink recipe for this project includes (with variations for different models):

-   Predict the target variable `target` with all other variables

-   Remove ID/ non-relevant variables like `application_order` and `application_mode`

-   Encode categorical predictors (for parametric models, not applicable for Naive Bayes)

    -   One-hot = TRUE (for tree-based models)

-   Filter out variables have have zero variance

-   Center & scale all predictors

### Looking at interactions for potential recipes:

Since my dataset isn't very large, I'm going to perform an EDA on my training dataset to see if there are any insights that can be generated to inform my recipe steps.

First, I'll look at a correlation matrix across my numeric variables:

```{r}
#| echo: false

library(tidyverse)
library(here)
load(here("results/drop_out_split.rda"))

drop_out_train |> 
  select(where(is.numeric))|> 
  cor(use = "pairwise.complete.obs") |> 
  ggcorrplot::ggcorrplot() +
  theme(text = element_text(size = 0.1))
```

Some of the main insights that can be generated are that there is a strong negative relationship between gdp and unemployment rate. However, I personally don't think that these are important variables to look at since they mostly pertain to the national economy rather than individual students (I might not include them in the recipe). There is also a pretty strong positive correlation between admission grade and previous qualification grade. This means that students who were doing well academically before college also tend to do well academically during college. The number of curricular units approved and curricular unit grades for a student from the first semester is also pretty positively correlated with the curricular units approved and curricular unit grades from the second semester. This suggests that there could be possible interactions between a student's first and second semester curricular units.

Next, I decided to look at parents' qualifications:

```{r}
#| echo: false

# mother's qualifications and father's qualifications
drop_out_train |> 
  ggplot(aes(x = mothers_qualification)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90))

drop_out_train |> 
  ggplot(aes(x = fathers_qualification)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90))
```

As we can see here, the distribution of qualifications for a student's mother and father are pretty similar. This suggests that there are possible interactions between these two variables.

Next, I decided to look at gender and whether or not it is correlated with the target variable:

```{r}
#| echo: false

# gender distribution within target
drop_out_train |> 
  ggplot(aes(x = target, fill = gender)) +
  geom_bar()
```

The bar graph shows that there are a couple of discrepancies in the gender proportion between these three target outcomes. It looks like a larger proportion of female students tend to graduate and are enrolled compared to male students. It also looks like the gender distribution for dropping out is pretty evenly split.

Next, I looked at variables that were indicative of socio-economic class:

```{r}
#| echo: false

# socio-economic class
drop_out_train |> 
  ggplot(aes(x = target, fill = debtor)) +
  geom_bar()

drop_out_train |> 
  ggplot(aes(x = target, fill = tuition_fees_up_to_date)) +
  geom_bar()

drop_out_train |> 
  ggplot(aes(x = target, fill = scholarship_holder)) +
  geom_bar()
```

Overall, it looks like a larger proportion of students who are debtors and don't have their tuition fees up to date tend to drop out compared to the students who are debtors and don't have their tuition fees up to date in the other two other outcome categories. It also looks like a larger proportion of students who are not scholarship holders tend to drop out compared to students who are scholarship holders.

After performing a short EDA, I plan on creating a recipe that removes unrelated variables such as unemployment rate, GDP, and inflation (wasn't really correlated with any other variables). A couple variables that would be important to look at are a student's gender, debtor status, whether or not their tuition is up to date, and whether or not they are a scholarship holder. I also plan on creating interaction variables between first semester curricular units and second semester curricular units, admission grade and previous qualification grade, and father and mother's qualifications. I also plan on adding a `step_corr` for my tree-based models to potentially remove variables that have large absolute correlations with other variables.

## Baseline Model:

For my baseline model, I decided to use a Naive Bayes model. I fit the model in the R Script `03_fit_naive_bayes`.

## Multinomial Model:

I'm also going to fit a Multinomial model with the baseline kitchen sink recipe to see if it performs better than the baseline Naive Bayes model. This model was fit in the R Script `03_fit_multinomial`.

## Comparing Naive Bayes to Multinomial model results:

```{r}
#| echo: false

load(here::here("results/baseline_results_table.rda"))

baseline_results_table |> 
  knitr::kable(col.names = c("Metric", "Mean", "Std. Err.", "Model"))
```

As we can see from the results table, the Multinomial model has a higher mean accuracy and ROC value compared to the Naive Bayes model. Although the Naive Bayes model has a slightly lower overall standard error across it's folds compared to the Multinomial model, these SE values are still very low for both models. The mean accuracy for the Multinomial model is around 0.75. This is a lot higher than the mean accuracy for the Naive Bayes, which is around 0.70. The mean ROC value for the Multinomial model is also better because it's closer to 1. Thus, the Multinomial model performed better than the Naive Bayes model.

## Progress Summary:

I was able to split my testing and training datastes, set up my folds, define a baseline kitchen sink recipe, and fit the recipe to my baselineline model and a Multinomial model. I currently have files set up to tune my other model types and I have a rough idea on what kind of recipe I would need to make going forwards.
